Shapes analysis for time series.

### 2024-09-04 NEURIPS24/paper.tex
update

With this in mind, the velocity field $v$ fitting the data can be
estimated by minimizing 
$v \in \rml^2([0,1],\msv) \mapsto \mathscr{L}(\phi^{v}.\mathbf{x},\mathbf{y})$, where $\mathscr{L}$ is an appropriate loss function.
 However, two computational challenges arise.
  First, this optimization problem is ill-posed, and a penalty term is needed to obtain a unique solution.
   In addition, we have to find a parametric family $\msv_{\Theta} \subset \rml^2([0,1],\msv)$, parameterized by $\Theta$, which allows us to solve this minimization problem efficiently.

%\paragraph*{Regularizing the problem to derive uniqueness.}
It has been proposed in \cite{miller2006geodesic} to interpret $\msv$ as a tangent space relative to the group of diffeomorphisms $\msh=\{ \phi^v:\eqsp v\in \rml^2([0,1],\msv)\}$.
Following this geometric point of view, geodesics can be constructed on $\msh$ by using the following squared norm 
 \begin{equation}
  \label{eq:geodesics_original}
    \mathscr{R}^2: g\in \msh\mapsto \inf_{ v\in \rml^2([0,1],\msv):\eqsp g=\phi^v} \int_0^1 || v_\tau||_\msv\dd \tau
 \end{equation}
By deriving differential constraints related to the minimum of \eqref{eq:geodesics_original} and using Cauchy-Lipschitz conditions, geodesics can be defined only by giving the starting point and the initial velocity $v_0\in \msv$ \cite{miller2006geodesic}, as straight lines in Euclidean space.
Denoting by $w(v_0)$ the geodesic starting from the identity with inital velocity $v_0$, the exponential map is defined as $\varphi^{\{v_0\}}\triangleq \phi^v$ and the previous matching problem becomes a \textit{geodesic shooting problem}:
 \begin{equation}
  \label{eq:geodesics_shooting}
  \inf_{v_0 \in \msv} \mathscr{L}(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}).
 \end{equation}
 Using $\varphi^{\{v_0\}}$ instead of $\phi^v$ for any $v\in \rml^2([0,1],\msv)$ regularizes the problem and induces a sparse representation for the learning diffeomorphisms.
 Moreover, by setting $\msv$ as an RKHS, the geodesic shooting problem has a unique solution and becomes tractable, as described in the next section.

@@

In this part, $\msv$ is chosen as an RKHS \cite{berlinet2011reproducing} generated by a smooth kernel $K$ (e.g., Gaussian). 
We follow \cite{durrleman2013sparse} and define a 
 discrete parameterization of the velocity fields to perform geodesics shooting \eqref{eq:geodesics_shooting}.  
  The initial velocity field $v_0$ is chosen as a finite linear combination of the RKHS basis vector fields, 
$\mathbf{n}_0$ control points $\msx_0=(x_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0}$ and momentum vectors $\alpha_0=(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0} $ are defined such that for any $x\in \Rset^{d'}$, 
  \begin{equation}
    \label{eq:def_v0}
    v_0\left(\alpha_0,\msx_0\right)(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_{k,0})\alpha_{k,0} \eqsp .
  \end{equation}
   In our applications, the control points $(x_{k,0})_{k\in[\mathbf{n}_0]}$ can be understood as the discretized graph $(t_k,\mathbf{s}_0(t_k))_{k\in[\mathbf{n}_0]}$ of a starting time series $\mathbf{s}_0$. 
  With this parametrization of $v_0$, \citet{miller2006geodesic} show that the velocity field $v$ of the solution of \eqref{eq:geodesics_shooting} keeps the same
  structure along time, such that for any $x\in\Rset^{d'}$ and $\tau\in[0,1]$, 
  \begin{equation}
    \label{eq:specific_form}
    v_\tau(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_k(\tau))\alpha_{k}(\tau) \eqsp ,
  \end{equation}
  \begin{equation} 
    \label{eq:integration}
      \left\{
        \begin{aligned}
        & \frac{\dd x_k(\tau)}{\dd \tau}=v_\tau(x_k(\tau)) \eqsp, \quad
        \frac{\dd \alpha_k(\tau)}{\dd \tau}=-\sum_{k=1}^{\mathbf{n}_0} \dd_{x_k(\tau)} K(x_k(\tau),x_l(\tau))\alpha_{l}(\tau)^\top \alpha_{k}(\tau) \eqsp  \\
        & \alpha_k(0)=\alpha_{k,0},\quad x_k(0)=x_{k,0} \eqsp , k\in[\mathbf{n}_0] 
        \end{aligned}
        \right .
  \end{equation}
  These equations are derived from the hamiltonian $H:(\alpha_k,x_k)_{k\in [\mathbf{n}_0]}\mapsto \sum_{k,l=1}^{\mathbf{n}_0} \alpha_{k}^\top K(x_k,x_l)\alpha_{l}  $, such that
  the velocity norm is preserved $||v_\tau||_{\msv}=||v_0||_\msv $ for any $\tau\in [0,1]$.
   By \eqref{eq:integration}, the velocity field related to a geodesic $v^*$ is fully parametrized by its initial control points and momentum $(x_{k,0},\alpha_{k,0})_{k\in[\mathbf{n}_0]}$.
   Thus, given a set of targets $\mathbf{y}=(y_i)_{i\in[T_2]}$ in $\Rset^{d'}$, a set of starting points $\mathbf{x}=(x_{i,0})_{i\in[T_1]}$ in $\Rset^{d'}$, a RKHS's kernel $K:\Rset^{d'}\times \Rset^{d'}\to \Rset^{d'\times d'}$, a distance on sets $\mathscr{L}$, 
 a numerical integration scheme of ODE and a penalty factor $\lambda>0$, the basic geodesic shooting step minimizes the following function using a gradient descent method:
   \begin{equation}
    \label{eq:relaxation}
    \mathcal{F}_{\mathbf{x},\mathbf{y}}: (\alpha_k)_{k\in [T_1]}\mapsto \mathscr{L}\left(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}\right)+\lambda||v_0||_\msv^2 \eqsp,  
   \end{equation}
   where $v_0$ is defined by \eqref{eq:def_v0} and $\varphi^{\{v_0\}}.\mathbf{x}$ is the result of the numerical integration of \eqref{eq:integration} using control points $\mathbf{x}$ and initial momentums $(\alpha_k)_{k\in[T_1]} $.

@@

@@

%\sam{améliorer transition}
\vspace{-1ex}
\paragraph{Loss}

This section specifies the distance function $\scrl$ introduced in the loss function defined in \eqref{eq:relaxation}.

>>>>

With this in mind, the velocity field $v$ fitting the data can be
estimated by minimizing 
$v \in \rml^2([0,1],\msv) \mapsto \mathscr{L}(\phi^{v}.\mathbf{x},\mathbf{y})$, where $\mathscr{L}$ is an appropriate loss function.
 However, two computational challenges arise.
  First, this optimization problem is ill-posed, and a penalty term is needed to obtain a unique solution.
   In addition, a parametric family $\msv_{\Theta} \subset \rml^2([0,1],\msv)$, parameterized by $\Theta$, is sought to efficiently solve this minimization problem. 
\paragraph*{From deformations to geodesics.}
%\paragraph*{Regularizing the problem to derive uniqueness.}
It has been proposed in \cite{miller2006geodesic} to interpret $\msv$ as a tangent space relative to the group of diffeomorphisms $\msh=\{ \phi^v:\eqsp v\in \rml^2([0,1],\msv)\}$.
Following this geometric point of view, geodesics can be constructed on $\msh$ by using the following squared norm 
 \begin{equation}
  \label{eq:geodesics_original}
    \mathscr{R}^2: g\in \msh\mapsto \inf_{ v\in \rml^2([0,1],\msv):\eqsp g=\phi^v} \int_0^1 || v_\tau||_\msv\dd \tau
 \end{equation}
By deriving differential constraints related to the minimum of \eqref{eq:geodesics_original} and using Cauchy-Lipschitz conditions, geodesics can be defined only by giving the starting point and the initial velocity $v_0\in \msv$ \cite{miller2006geodesic}, as straight lines in Euclidean space.
Denoting by $\tau \mapsto \rho_{v_0}(\tau)\in\msh$ the geodesic starting from the identity with inital velocity $v_0\in \msv$, the exponential map is defined as $\varphi^{\{v_0\}}\triangleq \rho_{v_0}(1)$. Using $\varphi^{\{v_0\}}$ instead of $\phi^v$, the previous matching problem becomes a \textit{geodesic shooting problem}:
 \begin{equation}
  \label{eq:geodesics_shooting}
  \inf_{v_0 \in \msv} \mathscr{L}(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}).
 \end{equation}
 Using $\varphi^{\{v_0\}}$ instead of $\phi^v$ for any $v\in \rml^2([0,1],\msv)$ regularizes the problem and induces a sparse representation for the learning diffeomorphisms.
 Moreover, by setting $\msv$ as an RKHS, the geodesic shooting problem has a unique solution and becomes tractable, as described in the next section.

@@

In this part, $\msv$ is chosen as an RKHS \cite{berlinet2011reproducing} generated by a smooth kernel $K$ (e.g., Gaussian). 
We follow \cite{durrleman2013sparse} and define a 
 discrete parameterization of the velocity fields to perform geodesics shooting \eqref{eq:geodesics_shooting}.  
  The initial velocity field $v_0$ is chosen as a finite linear combination of the RKHS basis vector fields, 
$\mathbf{n}_0$ control points $\msx_0=(x_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0}$ and momentum vectors $\alpha_0=(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0} $ are defined such that for any $x\in \Rset^{d'}$, 
  \begin{equation}
    \label{eq:def_v0}
    v_0\left(\alpha_0,\msx_0\right)(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_{k,0})\alpha_{k,0} \eqsp .
  \end{equation}
   In our applications, the control points $(x_{k,0})_{k\in[\mathbf{n}_0]}$ can be understood as the discretized graph $(t_k,\mathbf{s}_0(t_k))_{k\in[\mathbf{n}_0]}$ of a starting time series $\mathbf{s}_0$. 
  With this parametrization of $v_0$, \cite{miller2006geodesic} show that the velocity field $v$ of the solution of \eqref{eq:geodesics_shooting} keeps the same
  structure along time, such that for any $x\in\Rset^{d'}$ and $\tau\in[0,1]$, 
  \begin{equation}
    \label{eq:specific_form}
    v_\tau(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_k(\tau))\alpha_{k}(\tau) \eqsp ,
  \end{equation}
  \begin{equation} 
    \label{eq:integration}
      \left\{
        \begin{aligned}
        & \frac{\dd x_k(\tau)}{\dd \tau}=v_\tau(x_k(\tau)) \eqsp, \quad
        \frac{\dd \alpha_k(\tau)}{\dd \tau}=-\sum_{k=1}^{\mathbf{n}_0} \dd_{x_k(\tau)} K(x_k(\tau),x_l(\tau))\alpha_{l}(\tau)^\top \alpha_{k}(\tau) \eqsp  \\
        & \alpha_k(0)=\alpha_{k,0},\quad x_k(0)=x_{k,0} \eqsp , k\in[\mathbf{n}_0] 
        \end{aligned}
        \right .
  \end{equation}
  These equations are derived from the hamiltonian $H:(\alpha_k,x_k)_{k\in [\mathbf{n}_0]}\mapsto \sum_{k,l=1}^{\mathbf{n}_0} \alpha_{k}^\top K(x_k,x_l)\alpha_{l}  $, such that
  the velocity norm is preserved $||v_\tau||_{\msv}=||v_0||_\msv $ for any $\tau\in [0,1]$.
   By \eqref{eq:integration}, the velocity field related to a geodesic $v^*$ is fully parametrized by its initial control points and momentum $(x_{k,0},\alpha_{k,0})_{k\in[\mathbf{n}_0]}$.
   Thus, given a set of targets $\mathbf{y}=(y_i)_{i\in[T_2]}$ in $\Rset^{d'}$, a set of starting points $\mathbf{x}=(x_{i,0})_{i\in[T_1]}$ in $\Rset^{d'}$, a RKHS's kernel $K:\Rset^{d'}\times \Rset^{d'}\to \Rset^{d'\times d'}$, a distance on sets $\mathscr{L}$, 
 a numerical integration scheme of ODE and a penalty factor $\lambda>0$, the basic geodesic shooting step minimizes the following function using a gradient descent method:
   \begin{equation}
    \label{eq:relaxation}
    \mathcal{F}_{\mathbf{x},\mathbf{y}}: (\alpha_k)_{k\in [T_1]}\mapsto \mathscr{L}\left(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}\right)+\lambda||v_0||_\msv^2 \eqsp,  
   \end{equation}
   where $v_0$ is defined by \eqref{eq:def_v0} and $\varphi^{\{v_0\}}.\mathbf{x}$ is the result of the numerical integration of \eqref{eq:integration} using control points $\mathbf{x}$ and initial momentums $(\alpha_k)_{k\in[T_1]} $.

@@

\sam{Lien avec functional data analysis ? Insister sur le fait que concaténer temps et espace c'est moisi}

@@

%\sam{améliorer transition}
\vspace{-1ex}
\paragraph{Loss}
\sam{Simplifier cette partie ?}
This section specifies the distance function $\scrl$ introduced in the loss function defined in \eqref{eq:relaxation}.
